Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex

We introduce an ARG inference method (schematic in Figure 1) that uses (1) an
efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
collection dates of the SARS-CoV-2 samples to determine the order of genome
copying; and (3) several parsimony-based heuristics to locally refine the
topology. In brief, this combined approach uses the likelihood model of LS to
attach sample genomes to existing nodes and to detect recombinants, but because
many of the tip samples attached daily harbour few, if any, mutational changes,
we use parsimony to resolve the relationships near the tip samples.

@@

The generative process of the LS model is summarised in Figure 2. Here, a
transition matrix, Q, governs the process of switching (recombining) between
members of the reference panel (the hidden states). An emission matrix,
\textit{E}, allows for differences between the focal sequence and the hidden
state from which it is copied (due to mutation as well as genotyping or
sequencing error). Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to
find the most likely copying path through the sequences, given a set of
reference sequences, \textit{Q}, and \textit{E}.

@@

In the wide ARG (sampled to mid-2021), all these sequences are descendants of
recombination nodes, except MILK-103C712, which was removed during
preprocessing (Table 2); additionally, all are HMM-consistent, except
QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
and (2) the breakpoint intervals (Table 2). In all cases, the two methods
proposed identical or very closely related parent lineages. The general mosaic
genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
are two switches in the HMM copying paths (indicative of two breakpoints). The
breakpoint intervals estimated using the two methods overlap substantially in
all the cases (Table 2). Within each of the groups A to D, the constituent
sequences have identical breakpoint intervals, because they trace back to a
single recombination node in the wide ARG. This result is consistent with the
hypothesis that the sequences of the groups A to D emerged from four separate
recombination events (Jackson et al. 2021).

@@

@@

\begin{table} \centering \begin{tabular}{l|c|c|c|c|c} \hline
\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline \textbf{Sample} &
\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} &
\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\ \hline ALDP-11CF93B & A
& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
\\ B.1.177} & 7,729–10,870 \\ \hline \end{tabular}
\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
variant reported by Jackson et al. \citep{Jackson2021-ik} have recombinant
ancestry in the wide ARG. The breakpoint intervals and Pango lineage
assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
al., except the Pango lineage assignment of the parents of group B
recombinants, which were taken from Table 1 (motif-based results).} \end{table}

\begin{figure} \centering
\includegraphics[width=0.8\textwidth]{figures/ls_diagram_covid.pdf}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
The LS model in its implementation as an HMM is governed by the transition and
emission matrices, \textit{Q} and \textit{E}, respectively. Each may be a
function of the reference panel members, but transitions are generally assumed
to be independent of the hidden states themselves (shown in the pink panel).
This dramatically increases performance as the state space drops to two states
(i.e., switching or not switching). The recombination probability at site
\textit{l} is \textit{rl}, while n denotes the number of members in the
reference panel. Emissions may also be a function of the nucleotide states, but
in our case, we assume that all observed alleles (in both the reference panel
and focal sequence, \textit{al}) at the site under consideration are equally
likely and occur with mutation probability \textit{l} (shown in the blue
panel). These two processes, transition and emission (which encode
recombination and mutation, respectively), define the generative process of the
HMM, by which an imperfect mosaic is emitted from a reference panel of samples
and nodes from the ARG inserted at earlier collection dates (shown in the lower
panel). Here, the unseen states in the reference panel are shown as coloured
lines enclosed by the grey box, and the true path through the data which leads
to the emitted focal sequence is displayed by the black arrow. Examples of
transition and emission probabilities along this trajectory are shown by the
red and blue arrows, respectively. Black crosses along the focal sequence show
genotyping errors or mutations. Note that in practice, we do not know the
pattern of colours along the focal sequence and must infer it using the Viterbi
algorithm.} \end{figure}

\begin{figure} \centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
\citep{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
derived state). The percentages of each mutation type from the wide ARG are
represented by blue bars and the percentages from Yi et al. by orange bars,
with the darker colours representing one direction (e.g., C > U) and the
lighter colours the reverse (e.g., U > C).} \end{figure}

>>>>

We introduce an ARG inference method, summarised in
Figure~\ref{fig:overview_sc2ts}.
% Note going to rewrite this anyway, so not trying to make it flow.
It uses (1) an
efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
collection dates of the SARS-CoV-2 samples to determine the order of genome
copying; and (3) several parsimony-based heuristics to locally refine the
topology. In brief, this combined approach uses the likelihood model of LS to
attach sample genomes to existing nodes and to detect recombinants, but because
many of the tip samples attached daily harbour few, if any, mutational changes,
we use parsimony to resolve the relationships near the tip samples.

@@

\begin{figure} \centering
\includegraphics[width=0.6\textwidth]{figures/ls_diagram_covid.pdf}
% FIXME this caption is too long
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
The LS model in its implementation as an HMM is governed by the transition and
emission matrices, $Q$ and $E$, respectively. Each may be a
function of the reference panel members, but transitions are generally assumed
to be independent of the hidden states themselves (shown in the pink panel).
This dramatically increases performance as the state space drops to two states
(i.e., switching or not switching). The recombination probability at site
$\ell$ is $r\ell$, while n denotes the number of members in the
reference panel. Emissions may also be a function of the nucleotide states, but
in our case, we assume that all observed alleles (in both the reference panel
and focal sequence, $a\ell$) at the site under consideration are equally
likely and occur with mutation probability $\ell$ (shown in the blue
panel). These two processes, transition and emission (which encode
recombination and mutation, respectively), define the generative process of the
HMM, by which an imperfect mosaic is emitted from a reference panel of samples
and nodes from the ARG inserted at earlier collection dates (shown in the lower
panel). Here, the unseen states in the reference panel are shown as coloured
lines enclosed by the grey box, and the true path through the data which leads
to the emitted focal sequence is displayed by the black arrow. Examples of
transition and emission probabilities along this trajectory are shown by the
red and blue arrows, respectively. Black crosses along the focal sequence show
genotyping errors or mutations. Note that in practice, we do not know the
pattern of colours along the focal sequence and must infer it using the Viterbi
algorithm.} \end{figure}

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}.
Here, a
transition matrix, $Q$, governs the process of switching (recombining) between
members of the reference panel (the hidden states). An emission matrix,
$E$, allows for differences between the focal sequence and the hidden
state from which it is copied (due to mutation as well as genotyping or
sequencing error). Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to
find the most likely copying path through the sequences, given a set of
reference sequences, $Q$, and $E$.

@@

\begin{table} \centering
\begin{tabular}{l|c|c|c|c|c} \hline
\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
\textbf{Sample} &
\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
interval(s)} &
\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
\hline ALDP-11CF93B & A
& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
\\ B.1.177} & 7,729–10,870 \\ \hline
\end{tabular}
\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
variant reported by \cite{Jackson2021-ik} have recombinant
ancestry in the wide ARG. The breakpoint intervals and Pango lineage
assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
al., except the Pango lineage assignment of the parents of group B
recombinants, which were taken from Table 1 (motif-based results).} \end{table}

In the wide ARG (sampled to mid-2021), all these sequences are descendants of
recombination nodes, except MILK-103C712, which was removed during
preprocessing (Table 2); additionally, all are HMM-consistent, except
QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
and (2) the breakpoint intervals (Table 2). In all cases, the two methods
proposed identical or very closely related parent lineages. The general mosaic
genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
are two switches in the HMM copying paths (indicative of two breakpoints). The
breakpoint intervals estimated using the two methods overlap substantially in
all the cases (Table 2). Within each of the groups A to D, the constituent
sequences have identical breakpoint intervals, because they trace back to a
single recombination node in the wide ARG. This result is consistent with the
hypothesis that the sequences of the groups A to D emerged from four separate
recombination events~\cite{Jackson2021-ik}.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra}
Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
Mutations are categorised by type (i.e., ancestral state >
derived state). The percentages of each mutation type from the wide ARG are
represented by blue bars and the percentages from Yi et al. by orange bars,
with the darker colours representing one direction (e.g., C > U) and the
lighter colours the reverse (e.g., U > C).}
\end{figure}

@@
